Updating to XCMS3 ...new way to organize the data which impacts all functions in XCMS

Nice tutorial on the new XCMS3: https://jotsetung.github.io/metabolomics2018/xcms-preprocessing-ioslides.html#1

Notes: 
* Rewrote peakShape.R function, with some final help from Johannes Rainer
* The code in the plotTIC chunk replaces the GetTICs_KL.r function 

Krista Longnecker
November 5, 2018
May 8, 2019 (cleaning up to work with IS)
========================================================

Go ahead and load up some of the parameters:
```{r Settings, eval=TRUE}
suppressMessages(library(xcms))
suppressMessages(library(CAMERA))
library(RColorBrewer)
library(XLConnect)
library(pander)
library(magrittr)

#Need these for plotting
library(ggplot2)
library(ggdendro)

#setup the option to use parallel processing
register(bpstart(SnowParam()))

```


Now go ahead and do the peak picking, or load in the existing file...peak picking can be time consuming and it easier to only do that once

```{r settingUpFiles, eval = TRUE}

#same file organization as in XCMS < 3
#path to the mzML files (need to set on each computer)

usePath <- "/data"

ext <- ".mzML"
pre <- paste0(usePath,"/")

mzdatafiles <- list.files(usePath,recursive = FALSE, full.names=TRUE,
                          pattern = paste0("\\",ext))

```


```{r getSampleInformation, eval=TRUE}
#Setup the ability to rename the classes based on information in the CSV file. 

#10/2018: setup the groups now...no need to do peak picking on data files that 
# I will not use (and easier with the new XCMS3 code since sampclass no longer exists in XCMS3)

wDir <-  "/ms/"

all <- readWorksheetFromFile(paste0(wDir,"simple_sampleList.xlsx"),sheet = 1)
putDataHere <-data.frame()

#what data do I want to keep...put that here
keep = "negative.1"

for(i in 1:length(all$File.Name)) {
  h <- all$File.Name[i]
  #paste0 will strip out the spaces; make the full filename here    
  h <-paste0(pre,h,ext)
  
  #pull tData now, so I can look for a match with the 'keep' variable I set at line 63
  tData <- all[i,]
  test = paste0(tData$ionMode,".",tData$goodData)
  
  #only proceed if test == keep
  tm = match(test,keep)
  
  if (!is.na(tm)) {
  #this will pick up the right row in mzdatafiles
    m <- match(basename(h),basename(mzdatafiles))
    if (!is.na(m)) {
      
      if (nrow(putDataHere)==0) {
        putDataHere <- tData
        putDataHere$FileWithExtension <- basename(h) #this will have the mzML extension...useful later
        
        } else {
          useIdx = nrow(putDataHere)+1
          putDataHere[useIdx,] <-tData
          putDataHere$FileWithExtension[useIdx] <- basename(h)
          rm(useIdx)
        }
    }
    rm(m)
  } #this ends the final if statement
    rm(h,test,tData,tm)
  
} #this ends the for loop

rm(all) #keep this clean...

```


``` {r readDataFiles}

#reading in the data files will be fast

my_data <- readMSData(files = paste0(pre,putDataHere$FileWithExtension),
                      pdata = new("NAnnotatedDataFrame",putDataHere),mode="onDisk")
#note that we are reading in data that has been converted to centroid mode using msconvert. The new XCMS3 can also do the centroid conversion, though I haven't tested it out.

# #can access the sample information with this code:
# pData(my_data)
# my_data$FileWithExtension #example of how to get individual variables

```



```{r settingFileNames}
# #setup the filenames all in one place bc easier...

#comment out the more complex version used with full datasets
# nDate <- "2019.05.06_v1"
# nBase <- "Weber_PPL_UPLC_neg_samples_KLworking"
# fName_allPeaks <- paste0(nBase,'_listAllPeaks.',nDate,'.csv')
# fName_aligned <- paste0(nBase,'_aligned.',nDate,'.csv')
# fName_peaksPicked <- paste0(nBase, '_picked.',nDate, '.pdf')
# #fName_TICplot <- paste0(nBase,'_TICplot.',nDate,'.pdf')
# NameOfFile <- paste0(nBase,"_",nDate,".Rimage")
# rm(nDate,nBase)

nBase <- "/output/simple_XCMS3_output"
fName_allPeaks <- paste0(nBase,'_listAllPeaks','.csv')
fName_aligned <- paste0(nBase,'_aligned','.csv')
fName_peaksPicked <- paste0(nBase, '_picked', '.pdf')
NameOfFile <- paste0(nBase,".Rimage")
rm(nDate,nBase)


save.image(NameOfFile)


```


```{r plotTIC, eval = TRUE}
## Get and plot the total ion chromatograph. This reads data from the files. 
#Note that this replaces the r function 'getTICs_KL.r' used with previous version of XCMS

tic <- chromatogram(my_data, aggregationFun = "sum")
## Define colors for the two groups (this variable gets used in later plotting)
group_colors <- brewer.pal(3, "Set1")[1:2]

#this needs to match the group names
names(group_colors) <- c("pooled","Unknown") 

## Plot all chromatograms.
plot(tic, col = group_colors[my_data$sampleDetails])

```

```{r plot_boxplot, eval = TRUE}

#first, pull the tic data from my_data; note that the 'tic' function uses the header information
# from the raw data files that I have read in
tc <- split(tic(my_data), f = fromFile(my_data))

#plot.new()
boxplot(tc, col = group_colors[my_data$sampleDetails],
    ylab = "intensity", main = "Total ion current") 

```

```{r plot_oneEIC, eval = TRUE}
#this will plot individual EICs in various ways, most of which seem useful [a series of plots after each iteration of plot.new() ]

rtr <- c(260, 270)
mzr <- c(245, 246) #using a window for d2-biotin
## extract the chromatogram
chr_raw <- chromatogram(my_data, mz = mzr, rt = rtr)
plot.new()
plot(chr_raw, col = group_colors[chr_raw$sampleDetails]) 

##not sure I understand this syntax...but it's in the vignette and it makes a nice set of figures
#for a given mzRT feature. Here are the notes from the vignette:
# For the ppm parameter we extract the full MS data (intensity, retention time and m/z values) corresponding to the above peak. To this end we first filter the raw object by retention time, then by m/z and finally plot the object with type = "XIC" to produce the plot below. We use the pipe (%>%) command better illustrate the corresponding workflow.
plot.new()
my_data %>%
    filterRt(rt = rtr) %>%
    filterMz(mz = mzr) %>%
    plot(type = "XIC") 

```

```{r, stilllMorePlotting, eval = TRUE}

#or, another example of plotting (Chromatogram extrcts the chromatographic data)
plot.new()
my_data %>%
    filterRt(rt = c(260, 270)) %>%
    filterMz(mz = c(245, 246)) %>%
    chromatogram(aggregationFun = "max") %>%
    plot()

#and still another figure, uses the data for d2-biotin
plot.new()
plot(chr_raw, col = group_colors[chr_raw$sampleDetails], lwd = 2)

```


```{r doPeakPicking, eval = TRUE}

setNoise = 10000 # Lumos noise levels are higher than that FT

#first setup the parameters for centWave
cwp <- CentWaveParam(peakwidth = c(3, 15), noise = setNoise, ppm = 15, mzCenterFun = "wMean",
                     prefilter = c(4,setNoise),integrate = 2, mzdiff = -0.005, 
                     fitgauss = TRUE, snthresh = 100)

#then use these parameters to find the peaks
xs <- findChromPeaks(my_data, param = cwp) 

```


```{r putPeakShapeHere, eval = TRUE, warning = FALSE, message = FALSE}
#KL re-wrote the peakShape.R function bc it relied heavily on the format of 
#data in the old XCMS (XCMS < 3.0)

source("peakShape_XCMS3.r")

if (1) {
  #run peakshape
  xdata <- peakShape_XCMS3(xs,cor.val = 0.9,useNoise = setNoise)
  
} else {
  #skip peakshape for now
  xdata <- xs  
  
}

```



```{r perFileSummary, eval =  TRUE}
#we use the data from this table to calculate some per-file summaries. 
#code from vignette

summary_fun <- function(z) {
    c(peak_count = nrow(z), rt = quantile(z[, "rtmax"] - z[, "rtmin"]))
}

T <- lapply(split.data.frame(chromPeaks(xdata),
                 f = chromPeaks(xdata)[, "sample"]),
        FUN = summary_fun)
T <- do.call(rbind, T)
rownames(T) <- basename(fileNames(xdata))
pandoc.table(T,
         caption = paste0("Summary statistics on identified chromatographic",
                  " peaks. Shown are number of identified peaks per",
                  " sample and widths/duration of chromatographic ",
                  "peaks.")) 

```


```{r plot_mzRTspace, eval = TRUE}

#eg, this will plot the data from file #3 in mz/rt space (plots only 1 file)
plotChromPeaks(xdata, file = 1) 

```


```{r plot_densityByColor, eval = TRUE}
# To get a global overview of the peak detection we can plot the frequency of identified peaks per file along the retention time axis. This allows to identify time periods along the MS run in which a higher number of peaks was identified and evaluate whether this is consistent across files.
#higher frequency is color code as yellow-white
#bin size is how much to divide up the figure (in seconds of retention time)
plot.new()
plotChromPeakImage(xdata, binSize = 10) 

##table for data in the figure, again specific for d2-biotin peak, but the table provides the details for one mz value
# Note that we can also specifically extract identified chromatographic peaks for a selected region by providing the respective m/z and retention time ranges with the mz and rt arguments in the chromPeaks method.
#note that mzr and rtr are set above, correspond to d2 biotin
pander(chromPeaks(xdata, mz = mzr, rt = rtr),
       caption = paste("Identified chromatographic peaks in a selected ",
               "m/z and retention time range.")) 

# Finally we plot also the distribution of peak intensity per sample. This allows to investigate whether systematic differences in peak signals between samples are present.

## Extract a list of per-sample peak intensities (in log2 scale)
# similar to code in plot_boxplots given above...
ints <- split(log2(chromPeaks(xdata)[, "into"]),
          f = chromPeaks(xdata)[, "sample"])
plot.new()
boxplot(ints, varwidth = TRUE, col = group_colors[xdata$sampleDetails],
    ylab = expression(log[2]~intensity), main = "Peak intensities")
#this next line draws dotted lines along the peak values
grid(nx = NA, ny = NULL)

```



```{r setUpAlignment, eval = TRUE}

#set up a new variable with the adjusted RT (will allow me to go back to xdata if needed)
rc.obi <- adjustRtime(xdata, param = ObiwarpParam(binSize = 0.1,distFun = "cor",
                                                 gapInit = 0.3, gapExtend = 2.4), msLevel = 1L)

#now plot up the results after retention time correction
## Get the base peak chromatograms.
bpis_adj <- chromatogram(rc.obi, aggregationFun = "max")
par(mfrow = c(2, 1), mar = c(4.5, 4.2, 1, 0.5))
plot(bpis_adj, col = group_colors[bpis_adj$sampleDetails])
## Plot also the difference of adjusted to raw retention time.
plotAdjustedRtime(rc.obi, col = group_colors[rc.obi$sampleDetails]) 

#Use adjustedRtime parameter to access raw/adjusted retention times
#again, using parameters set above for d2 biotin
par(mfrow = c(1, 2), mar = c(4, 4.5, 0.9, 0.5))
plot(chromatogram(rc.obi, mz = mzr,
                  rt = rtr, adjustedRtime = FALSE))
plot(chromatogram(rc.obi, mz = mzr,
                  rt = rtr)) 

## Does the object have adjusted retention times?
hasAdjustedRtime(rc.obi) #answer will be TRUE or FALSE

# ## Drop the alignment results (not clear why you want to do this...but here's the code)
# xdata <- dropAdjustedRtime(rc.obi)
# 
#if you want to replace the RAW retention times, use applyAdjustedRtime (KL note...this seems like a bad idea, so I am not including it here...)


```


```{r doGroupingNow, eval = TRUE}
# new name in XCMS3: 'correspondence'

# The final step in the metabolomics preprocessing is the correspondence that matches detected chromatographic peaks between samples (and depending on the settings, also within samples if they are adjacent). The method to perform the correspondence in xcms is groupChromPeaks. We will use the peak density method to group chromatographic peaks. The algorithm combines chromatographic peaks depending on the density of peaks along the retention time axis within small slices along the mz dimension. The sample group assignment is specified with the sampleGroups argument.

# The definition of the sample groups (i.e. assignment of individual samples to the sample groups in the experiment) is mandatory for the PeakDensityParam. If there are no sample groups in the experiment sampleGroups should be set to a single value for each file (e.g. rep(1, length(fileNames(xdata))).

## Correspondence: group peaks across sample using 'peak density'
#one definition for bw (from Jan Stanstrup): bw is the maximum expected RT deviation across samples
#sampleGroups = rc.obi$sampleDetails

pdp <- PeakDensityParam(sampleGroups = rc.obi$sampleDetails,
            minFraction = 0.1, minSamples = 1, bw = 3)
xgC <- groupChromPeaks(rc.obi, param = pdp)

#other option for correspondence is 'nearest', but that has not worked well for WHOI data

##testing [HERE]
# ## Now the retention time correction (this time using peakGroups, different option)
#pgp <- PeakGroupsParam(minFraction = 0.1)
# ## Get the peak groups that would be used for alignment.
#xRT <- adjustRtime(xgC, param = pgp)

# Note also that we could use the adjustedRtime / PeakGroups method on the object before alignment to evaluate on which features (peak groups) the alignment would be performed. 

# ## Plot the difference of adjusted to raw retention time.
# plotAdjustedRtime(xgC, col = group_colors[xgC$sampleDetails],
#           peakGroupsCol = "grey", peakGroupsPch = 1) 

#At last we evaluate the impact of the alignment on the test peak (d2 biotin).
par(mfrow = c(2, 1))
## Plot the raw data
plot(chr_raw, col = group_colors[chr_raw$sampleDetails])

## Extract the chromatogram from the adjusted object
chr_adj <- chromatogram(xgC, rt = rtr, mz = mzr)
plot(chr_adj, col = group_colors[chr_raw$sampleDetails]) 

```

```{r plot_peakDensityAXsamples, eval = TRUE}
# plot the chromatogram for an mz slice with multiple chromatographic peaks within each sample. 

#default values for now

## use mzr and rtr defined above for d2-biotin

## Extract and plot the chromatograms
chr_mzr <- chromatogram(xgC, mz = mzr, rt = rtr)
par(mfrow = c(3, 1), mar = c(1, 4, 1, 0.5))
cols <- group_colors[chr_mzr$sampleDetails]
plot(chr_mzr, col = cols, xaxt = "n", xlab = "")
## Highlight the detected peaks in that region.
highlightChromPeaks(xgC, mz = mzr, col = cols, type = "point", pch = 16)


#will plot multiple windows
par(mar = c(4, 4, 1, 0.5))

## Define the parameters for the peak density method. In this figure, 
    # Points are peaks per sample;
    # black line: peak density distribution;
    # grey rectangles: grouped peaks (features).

pdp <- PeakDensityParam(sampleGroups = xgC$sampleDetails,
            minFraction = 0.4, bw = 3)
plotChromPeakDensity(xgC, mz = mzr, col = cols, param = pdp,
             pch = 16, xlim = c(rtr[1] - 50, rtr[2] + 50))

## Use a different bw
pdp <- PeakDensityParam(sampleGroups = xgC$sampleDetails,
            minFraction = 0.4, bw = 8)
plotChromPeakDensity(xgC, mz = mzr, col = cols, param = pdp,
             pch = 16, xlim = c(rtr[1] - 50, rtr[2] + 50)) 

# The results from the correspondence can be extracted using the featureDefinitions method, that returns a DataFrame with the definition of the features (i.e. the mz and rt ranges and, in column peakidx, the index of the chromatographic peaks in the chromPeaks matrix for each feature).

## Extract the feature definitions
featureDefinitions(xgC)

# The featureValues method returns a matrix with rows being features and columns samples. The content of this matrix can be defined using the value argument. Setting value = "into" returns a matrix with the integrated signal of the peaks corresponding to a feature in a sample. Any column name of the chromPeaks matrix can be passed to the argument value. Below we extract the integrated peak intensity per feature/sample.

## Extract the into column for each feature.
head(featureValues(xgC, value = "into"))

```


```{r doFillPeaks, eval = TRUE}
## Filling missing peaks using default settings. Alternatively we could
## pass a FillChromPeaksParam object to the method.
save.image(NameOfFile)

fillParam <- FillChromPeaksParam(expandMz = 0, expandRt = 0, ppm = 0)
xgF <- fillChromPeaks(xgC,fillParam)

head(featureValues(xgF))

# Below we compare the number of missing values before and after filling in missing values. We can use the parameter filled of the featureValues method to define whether or not filled-in peak values should be returned too.

## Missing values before filling in peaks (set by filled = FALSE)
apply(featureValues(xgF, filled = FALSE), MARGIN = 2,
      FUN = function(z) sum(is.na(z)))

## Missing values after filling in peaks
apply(featureValues(xgF), MARGIN = 2,
      FUN = function(z) sum(is.na(z)))

```



```{r getHistory, eval = TRUE}
# XCMSnExp objects allow to capture all performed pre-processing steps along with the used parameter class within the @processHistory slot. Storing also the parameter class ensures the highest possible degree of analysis documentation and in future might enable to replay analyses or parts of it. The list of all performed preprocessings can be extracted using the processHistory method.

processHistory(xgF) 

# It is also possible to extract specific processing steps by specifying its type. Available types can be listed with the processHistoryTypes function. Below we extract the parameter class for the alignment/retention time adjustment step.

ph <- processHistory(xgF, type = "Retention time correction")

#And we can also extract the parameter class used in this preprocessing step.

## Access one parameter
processParam(ph[[1]])

```


Go ahead and spit out some variables to the folder

```{r SaveData, eval = TRUE}

#this will spit out the list of all features identified
#previous version of XCMS : write.csv(xgF@peaks,file = fName_allPeaks)
#version for XCMS3; this will also include a variable 'is_filled' to let the user know when a peak
#was found using fillPeaks
allPeaks <- chromPeaks(xgF)
write.csv(allPeaks, file = fName_allPeaks)

```

These are the pieces for CAMERA
      
```{r CAMERA, eval = TRUE}

##CAMERA is not setup to work with the output from XCMS3...so have to change datafile to xcmsSet
xset_new <- as(xgF, "xcmsSet")

#not entirely clear to me why this next step has to be repeated, but something about new XCMS
#and old CAMERA
xset_new <- fillPeaks(xset_new)
#
#nSamples <- length(mzdatafiles)
#xsa<-xsAnnotate(xgF,sample = c(1:nSamples))
#xsa<-xsAnnotate(xgF,sample = 34) randomly try one sample for now
xsa<-xsAnnotate(xset_new)

#group the features initially just by retention time
xsaF <-groupFWHM(xsa)

nSamples <- length(mzdatafiles)
#figure out which features also have a matching 13C feature. Have to enter both 
#the relative error (ppm) and the absolute error (mzabs)
xsaFI <-findIsotopes(xsaF,ppm=3,mzabs = 0.01,minfrac = 1/nSamples,intval = c("into"))

#now group by the correlations based on (1) intensity, (2) EIC, (3) isotopes...
xsaC <-groupCorr(xsaFI,cor_eic_th=0.75,pval=0.05, graphMethod="hcs",
 calcIso = TRUE, calcCiS = TRUE, calcCaS = FALSE)

#setup the file to also look for adducts, only go with the primary adducts for the moment
file <-system.file("rules/primary_adducts_neg.csv",package = "CAMERA")
rules <-read.csv(file)
an <-findAdducts(xsaC,polarity = "negative",rules=rules,ppm=3)

#do some housecleaning
rm(xsa,xsaF,xsaFI,xsaC)
```


```{r makePDFofFeatures_new, eval = TRUE}
#extract the EICs from the grouped XCMSnExp object:
chrs <- featureChromatograms(xgF)

## Extract also XICs without adjusted retention times
chrs_raw <- featureChromatograms(xgF, adjustedRtime = FALSE)

# ## (example) Plot the XIC for the first feature using different colors for each file
# par(mfrow = c(1, 2))
# plot(chrs[1, ], col = c("red", "green", "blue"))
# plot(chrs_raw[1, ], col = c("red", "green", "blue"))

# I want to loop over nrow(chrs) to plot the data for each feature:

#preset the colors, use ncol to figure out how many samples I have (and hence # of colors needed)
nColors <- ncol(chrs)
useColors <- brewer.pal(max(nColors,3),'Dark2')

cairo_pdf(file = fName_peaksPicked,onefile=TRUE)
#for (i in 1:nrow(chrs)){
for (i in 1:100) {  ##short for testing
    par(mfrow = c(3,1)) #set to c(3,1) if TRUE in following if statement, otherwise c(2,1)
    plot(chrs[i, ], col = useColors)
    plot(chrs_raw[i, ], col = useColors)
    
    # Can also use 'highlightChromPeaks' to indicate the identified 
    #peaks in the individual XICs (change IF statement to TRUE:
    if (TRUE) {
      plot(chrs[i, ], col = useColors)
      highlightChromPeaks(xgF, rt = range(lapply(chrs[i, ], rtime)), 
                        mz = range(lapply(chrs[i, ], mz)),
                        border = useColors)
    }
 }
dev.off()


```

And save the final steps

```{r lastSave, eval = TRUE}
##annoying, but CAMERA is not yet ported to using the data format from XCMS3, so need some hacking to ##have the data exported as I am used to ...this is entirely an issue where xsAnnotate is truncating the file names to just the final number (e.g. X12 for the file withSomeLongName_12)

#solution for now...get the peak list, correct the names, and then export the list to a CSV file
tExport <- getPeaklist(an)
new_columns <- colnames(tExport)

#get full filenames...
fullFilenames <- an@xcmsSet@phenoData$File.Name

#make life easy, pull the short names...used code from Craig to follow to this idea
shortFilename <- list()
for (i in 1:length(fullFilenames)) {
  w <- (unlist(strsplit(fullFilenames[i],"_")))
  getLast = w[[length(w)]]
  #put the annoying X at the beginning to match xsAnnotate output
  shortFilename[i] <- paste0("X",getLast)
  rm(w,getLast)
}

for (i in 1:length(new_columns)) {
  t <- new_columns[i]
  if (!is.na(pmatch("X",t))) {
    #go get the right file name and put it into columns
    index <- match(t,shortFilename)
    new_columns[i] <- fullFilenames[index]
    rm(index)
  }
  rm(t)
}
#now put the new column names into tExport
new_columns -> colnames(tExport)
write.csv(tExport, file = fName_aligned)
rm(new_columns,tExport,shortFilename, fullFilenames) #house cleaning

```

``` {r saveFinalImage, eval = TRUE}

save.image(NameOfFile)


```
